Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-12-02 response_to_reviewers.txt

We thank the reviewers for their suggestions.

We are especially grateful to Reviewer 2, who has offered the most detailed and constructive review we may have ever received.

Reviewer 1

I think that the abstract is too long and should be three-time shorter.
* The abstract is now substantially shorter. We have cut it down as much as we felt possible without losing clarity.

@@

Page 14

56-: The connection between a smaller κ and a right-shift of the access time distribution is difficult to see as differences in the saturation are hardly noticable.

Page 15

* General response to the following figure-related comments:
* Thank you for the comments and we agree that it is difficult to parse out the various dimensions shown here. However, we are faced with a rather difficult visualization problem that has many variables: access time, count, degrees, and curvature. After quite a bit of effort, this was the best that we could do. To make it possible for people such as this reviewer to dig a little more deeply into the data, we have now linked to an archived version of the exact data points used to make the plots, in their five-dimensional overlapping glory. We also note that the statements regarding the influence of curvature are confirmed by the regression analysis.

@@

Plots of Figure 5:
1. It is very hard to extract any information from the plots, because the plots (as soon as there are a few more lines) quickly get congested due to overlaps and hardly (or not at all) visible differences in saturation.

2. Adding axis labels (and not just explaining them in the figure caption) could help the readability of the plots.

**** TODO: Chris, could you add these axis labels with the same fonts?

@@

2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
* These plots demonstrate the effect of curvature on access times when variables such as degree are controlled. Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.  A pair of trees that both contain leaves 1 and 2 as siblings are an example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves. We rephrased this sentence to state "To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g. pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."

Page 17

12-15: Adding hyphens (e.g. 'distance-1 pairs') might increase the readability.
* We have made the suggested change

15-23: The traceability / comprehensibility of the statements would strongly benefit from improving the readability of the plots in Figure 5 and 6.
* See above, also noting that Tables 1 and 2 confirm several of the statements made in a very readable form.

### 2016-12-02 response_to_reviewers.txt

39:
1. There seems to be an inconsistency between the definition of internal edges and their number stated in this line. On page 5, internal edges are defined as those that do not contact a leaf or ρ. Then, a 'rooted tree with k leaves' cannot have 2(k-1) internal edges as stated here (because this is the number of all edges (also those leading to leaves), disregarding the extra root edge). So, either the definition of 'internal edge' has to be adopted or the word 'internal' has to be dropped here.
* We have dropped the word "internal" here.

2. The white space after 'edges' is missing.
* Corrected.

Page 20

9-19: While Corollary 5.4 appears correct, the given justification seems to be a bit problematic.  Based on the definition of γ and the excluded (subtracted) moves, it appears that these γ moves still include some which move R. For example, from the set of all possible moves for T (deg(T)), only some moves into R (2kb, case (ii)) and some moves of R's parent subtree (2(j-1), case (iv)) are excluded.  However, it is then stated that these moves 'can be applied to T and S with the same source and target nodes'. The minor (but arguable) problem concerns the part 'same source nodes': When are they the same? Is the parent node of R the same although its position differs in T and S? If not, then not all γ moves can be applied with the same source nodes. The bigger problem is created by the fact that moves moving R with the same target node eliminate the only difference between T and S and, thus, the difference between the obtained neighbours is zero (not one).  Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be obtained.
* We revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.".
* For the second part, we revised the proof of Corollary 5.4 to pair the set of shared neighbors with each other such that each pair of trees is at distance 1 rather than distance 0. As no optimal mass transport pairing would do this, we also added the following note before Lemma 5.6: "Note that these shared neighbors are exactly the trees from the second case of Corollary 5.4 and an optimal mass transport pairing will pair mass at these trees".

### 2016-12-02 curvature.tex
Addressing figure comments.
fix proof of cor:paired_neighbors

	\caption{\
        Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-leaf trees, grouped by the degree of the trees.
				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
	}
	\label{fig:kappa-access-distribution}
\end{figure*}

@@

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g. pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{cor}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that $\dspr{T',S'} = 1$ for each $(T',S')$ pair, each neighbor of $T$ is paired with a single neighbor of $S$, and vice versa.
Note that a tree may be a neighbor of both $T$ and $S$ and thus involved in two different pairings.
\end{cor}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
\end{proof}

@@

In our next lemma of this section, we bound the number of neighbors shared by two adjacent trees.
The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive, as mass assigned to a shared neighbor may not require a mass transport cost.
We show that the maximum number of shared neighbors is linear in the number of leaves.
In other words, only a small proportion of any two neighborhoods are shared and the relative size of shared neighborhoods grows much more slowly than the quadratic scaling of total neighborhood sizes.

>>>>

	\caption{\
        Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-leaf trees, grouped by the degree of the trees.
				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
    See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
	}
	\label{fig:kappa-access-distribution}
\end{figure*}

@@

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g. pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
        See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{cor}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that $\dspr{T',S'} = 1$ for each $(T',S')$ pair, each neighbor of $T$ is paired with a single neighbor of $S$, and vice versa.
Note that a tree may be a neighbor of both $T$ and $S$ and thus involved in two different pairings.
\end{cor}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
For each such $(T',S')$ pair, there are two cases, depending on whether $T'$ differs from $T$ by $R$ or another subtree.
If $T'$ and $T$ differ by a different subtree, then we can move $R$ in $T'$ or $S'$ to obtain the other member of the pair.
Thus, $\dspr{T',S'} = 1$.
If $T'$ and $T$ differ by $R$, however, then $T' = S'$.
Consider the set of such pairs $(T'_1, S'_1), (T'_2, S'_2), \ldots, (T'_p, S'_p)$.
Each tree of the pairs can be obtained from the others by moving $R$, so we simply pair $T'_1$ with $S'_2$, $T'_2$ with $S'_3$ and so on, finally pairing $T'_p$ with $S'_1$.
\end{proof}

@@

In our next lemma of this section, we bound the number of neighbors shared by two adjacent trees.
The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive, as mass assigned to a shared neighbor may not require a mass transport cost.
Note that these shared neighbors are exactly the trees from the second case of Corollary~\ref{cor:paired_neighbors} and an optimal mass transport pairing will pair mass at these trees.
We show that the maximum number of shared neighbors is linear in the number of leaves.
In other words, only a small proportion of any two neighborhoods are shared and the relative size of shared neighborhoods grows much more slowly than the quadratic scaling of total neighborhood sizes.
